Gravitational Lens Recovery with \Glass: How to measure the mass profile and shape of a lens

### 2013-08-02 limits/ms.tex
more edits

The arrival-time surface $\tau(\theta)$ is a theoretical construct for
understanding lensing observables, and not an observable in its own
right.

The only lensing observables from \eqnref{arrival time} are the image
positions. However, if the source object is variable then the observed
light curves of each image will be identical but shifted in time due
to the different photon path lengths.  The time difference between
identical portions of the light curves is called the time delay
$\Delta t$ between images.  Using \eqnref{tau} the time delay $\Delta
t_{21}$ between two images $\vec\theta_1$ and $\vec\theta_2$, where
$\vec\theta_2$ arrives after $\vec\theta_1$, is $\Delta t_{21} =
\Delta \tau_{21}(1+z_L)d_L / \zeta$. If $\beta$ and $\kappa_\infty$
could be known exactly this would allow one to precisely infer the
angular diameter distance to the lens and thus -- via a cosmological
model -- the Hubble constant. However, the source position is not
observable and the density distribution is highly degenerate.

Degeneracy first formally noted by \cite{1985ApJ...289L...1F}.  Most
important is mass sheet.  But in fact, there are infinitely many
\citep{2000AJ....120.1654S}.
Also shape degeneracies \citep{2006ApJ...653..936S,2013arXiv1306.4675S}
Suppressed by redshift contrast
\citep{1998AJ....116.1541A,2009ApJ...690..154S}
but not completely \citep{2008MNRAS.386..307L}.

\subsection{Degeneracies}\label{sec:degen} %-----------------------------------------------------

The trouble with solving the lens equation for the mass distribution is that the solutions are not unique. To see why, it is instructive to simplify \eqnref{eqn:lens_equation} for a circularly symmetric lens: 
%
\begin{equation}
\vec\beta = \vec\theta - \frac{\theta_E^2}{\theta^2}\vec\theta
\label{eqn:lens_equation_circ}
\end{equation}
%
where $\theta = |\vec\theta|$, $\theta_E^2 = \frac{D_{LS}}{D_S D_L} \frac{4 GM(<\theta)}{c^2}$ is the Einstein radius, and $M(<\theta)$ is the mass enclosed within the images.

For on-axis sources (\vec\beta\ = \vec0), the images split into a degenerate Einstein ring with $\theta = \theta_E$. Even in this situation of maximal information (since the source position $\vec\beta$ cannot typically be known), there is a trivial degeneracy between $M(<\theta_E)$ and the angular diameter distance. For fixed cosmology and in the most ideal situation, a single source splitting into images gives us only an enclosed mass. This `steepness' degeneracy (also known as the mass-sheet degeneracy) generalises to the non-circularly symmetric case \citep{1985ApJ...289L...1F,2000AJ....120.1654S}. Thus, to measure the mass profile of a lens, we require more information. This can come in the form of sources at different redshifts -- since each has its own Einstein radius, each constrains a different enclosed mass $M(<\theta_E')$. Or it can come in the form of additional constraints like time delay data between the images, stellar mass estimates from stellar populations synthesis modelling \citep{2008MNRAS.383..857F}, stellar kinematics \citep{2002MNRAS.337L...6T}, or similar. However, even with further information, higher order degeneracies remain. \citet{2008MNRAS.386..307L} showed that there is a monopole degeneracy that allows additional mass to be deposited in-between images without any optical effect, while \citep{2006ApJ...653..936S} find twisting/shape degeneracies that are sensitive to the time delays. Further degeneracies enter also if there is a strong {\it external} field acting on the lens -- for example from a larger group or cluster environment within which the lens sits \citep[e.g.][]{1996ApJ...468...17B}. However, this {\it shear term} can be constrained by data \citep[e.g.][]{2002ApJ...578...25K}. In \secref{sec:results}, we will use \Glass\ to determine how well such degeneracies can be broken by data of ever-increasing quality to measure the mass profile and shape of a lens. We assume throughout that the shear is negligible.

{\bf Is ``Schneider's degeneracy" something new? Should we cite it here? -- {\tt http://adsabs.harvard.edu/abs/2013arXiv1306.4675S}}

@@

\item Understand ``Schneider's degeneracy". Add reference/commentary accordingly.

>>>>

Lensing observations provide information only at $\vec\theta$ where
there are images.  Hence, the arrival-time surface $\tau(\vec\theta)$
is not itself observable.  Its usefulness lies in that observables can
be derived from it.  Thus an image observed at $\vec\theta_1$ implies
that $\nabla\tau(\vec\theta_1)=0$.  A measurement of time delays
between images at $\theta_1$ and $\theta_1$ implies that
$t(\vec\theta_1)-t(\vec\theta_2)$ is known.  Interestingly, both these
types of observations give contraints that are linear in $\kappa$ and
$\vec\beta$.

The rather complicated dependence of lensing observables on the mass
distribution $\kappa(\vec\theta)$ has an important consequence: very
different mass distributions can result in similar observables.  This
is the phenomenon of lensing degeneracies.  While the non-uniqueness
of lens models noted by \cite{1981ApJ...244..736Y} already hinted at
degeneracies, their existence was first derived by
\cite{1985ApJ...289L...1F}.  The most important is the so-called
mass-sheet degeneracy, which is that image positions remain invariant
if the $D_S/D_{LS} - \kappa(\vec\theta)$ are each multiplied by an
arbitrary constant.  This amounts to multiplying the arrival-time
surface by a constant factor.  In fact there are infinitely many
degeneracies \citep{2000AJ....120.1654S} because any transformation of
the arrival-time surface away from the images has no effect on the
lensing observables.  In particular, there are degeneracies that
involve the shape of the mass distribution
\citep{2006ApJ...653..936S,2013arXiv1306.4675S}.  Degeneracies tend to
be suppressed if there are sources at very different redshifts or
`redshift contrast' \citep{1998AJ....116.1541A,2009ApJ...690..154S},
because the presence of different factors of $D_S/D_{LS}$ in the image
plane makes it more difficult to change the mass distribution and the
arrival-time surface without affecting the lensing observables.  But
degeneracies are still present with multiple source redshifts.
\citep{2008MNRAS.386..307L}.

% The only lensing observables from \eqnref{arrival time} are the
% image positions. However, if the source object is variable then the
% observed light curves of each image will be identical but shifted in
% time due to the different photon path lengths.  The time difference
% between identical portions of the light curves is called the time
% delay $\Delta t$ between images.  Using \eqnref{tau} the time delay
% $\Delta t_{21}$ between two images $\vec\theta_1$ and
% $\vec\theta_2$, where $\vec\theta_2$ arrives after $\vec\theta_1$,
% is $\Delta t_{21} = \Delta \tau_{21}(1+z_L)d_L / \zeta$. If $\beta$
% and $\kappa_\infty$ could be known exactly this would allow one to
% precisely infer the angular diameter distance to the lens and thus
% -- via a cosmological model -- the Hubble constant. However, the
% source position is not observable and the density distribution is
% highly degenerate.

% \subsection{Degeneracies}\label{sec:degen}
%
% The trouble with solving the lens equation for the mass distribution is that the solutions are not unique. To see why, it is instructive to simplify \eqnref{eqn:lens_equation} for a circularly symmetric lens: 
%
% \begin{equation}
% \vec\beta = \vec\theta - \frac{\theta_E^2}{\theta^2}\vec\theta
% \label{eqn:lens_equation_circ}
% \end{equation}
%
% where $\theta = |\vec\theta|$, $\theta_E^2 = \frac{D_{LS}}{D_S D_L} \frac{4 GM(<\theta)}{c^2}$ is the Einstein radius, and $M(<\theta)$ is the mass enclosed within the images.

% For on-axis sources (\vec\beta\ = \vec0), the images split into a degenerate Einstein ring with $\theta = \theta_E$. Even in this situation of maximal information (since the source position $\vec\beta$ cannot typically be known), there is a trivial degeneracy between $M(<\theta_E)$ and the angular diameter distance. For fixed cosmology and in the most ideal situation, a single source splitting into images gives us only an enclosed mass. This `steepness' degeneracy (also known as the mass-sheet degeneracy) generalises to the non-circularly symmetric case \citep{1985ApJ...289L...1F,2000AJ....120.1654S}. Thus, to measure the mass profile of a lens, we require more information. This can come in the form of sources at different redshifts -- since each has its own Einstein radius, each constrains a different enclosed mass $M(<\theta_E')$. Or it can come in the form of additional constraints like time delay data between the images, stellar mass estimates from stellar populations synthesis modelling \citep{2008MNRAS.383..857F}, stellar kinematics \citep{2002MNRAS.337L...6T}, or similar. However, even with further information, higher order degeneracies remain. \citet{2008MNRAS.386..307L} showed that there is a monopole degeneracy that allows additional mass to be deposited in-between images without any optical effect, while \citep{2006ApJ...653..936S} find twisting/shape degeneracies that are sensitive to the time delays. Further degeneracies enter also if there is a strong {\it external} field acting on the lens -- for example from a larger group or cluster environment within which the lens sits \citep[e.g.][]{1996ApJ...468...17B}. However, this {\it shear term} can be constrained by data \citep[e.g.][]{2002ApJ...578...25K}. In \secref{sec:results}, we will use \Glass\ to determine how well such degeneracies can be broken by data of ever-increasing quality to measure the mass profile and shape of a lens. We assume throughout that the shear is negligible.

% {\bf Is ``Schneider's degeneracy" something new? Should we cite it here? -- {\tt http://adsabs.harvard.edu/abs/2013arXiv1306.4675S}}

@@
